Verifying Spatial Properties of Array Computations

### 2017-04-17 oopsla17/study.tex
edits to the study, intro and appendix

\item Many array computations are \emph{stencil computations}: an
  assignment whose right-hand side comprises array subscripts with
  neighbourhood indices and whose left-hand side is an
  neighbourhood-indexed array subscript (\eg{}, the Laplace operator);

@@

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we categorise the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l|l}
  Shorthand & Classification & Example \\ \hline
  (\textsf{const}) & Consistent & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  (\textsf{perm}) & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  (\textsf{LHSsub}) & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  (\textsf{LHSsup}) & LHS superset & \fortran{a(i, j) = b(i)} \\
  (\textsf{inconst}) & Inconsistent & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%

%\begin{itemize}
% (Vars, ...)
%\item An assignment to a variable, with an RHS comprising some array
%  computation, tends to correspond to a reduction (e.g., calculating
%the max value in an array).
%\end{itemize}

@@

\noindent
We revisit each hypothesis and show the related data and conclusions.
\begin{enumerate}
\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or dervied) induction variables;
We identified $108,773$ instances of assignments within loops in
which an array subscript flows to the right-hand side. We refer
to each one of these as an \emph{array computation}.
We performed the classification described above
and grouped the data by the right-hand side classification only:
\begin{tabular}{l|S[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
\textbf{RHS classification} & \textbf{Number} & \% &
\textit{grouping 1} \\ \hline
Affine                          & 34        & 0.03\%
& \multirow{3}{*}{81.43\%} \\ \cline{1-3}
Neighbourhood                   & 65334     & 60.06 \%  \\ \cline{1-3}
Neighbourhood + absolute       & 23209     & 21.34\%  \\ \hline
Other                           & 20196     & 18.57 \%  \\ \hline \hline
\textbf{Total}                  & 108773    &  \\
\noindent
Since, affine and neighbour patterns are static, we
group them (\textit{grouping 1}) to see that $81.43\%$ of the array
computations (88,577) have a static pattern that comprises
constant offsets of base/derived induction variables (potential with
some absolute indices), confirming
\emph{hypothesis 1}. Note that affine (derived induction variables)
are rare. Overwhelmingly, the main category is neighbourhood
offsets (possibly mixed with constants in some
dimensions). The ``affine + absolute'' class is
not represented at all.
\mnote{(1) State that constant here includes variables that are constant with
respect to the current loop scope.}
\mnote{(2) Mention what sort of things are
included in other.}
\mnote{(3) Add a remark about lack of affine indices to be
perhaps programmers relying on compiler more regarding optimisations than hand
optimising. We can later use this to justify our inability to handle affine
indices.}
\item Most loop-array computations of the previous form read
from arrays with a \emph{contiguous} pattern, \eg{}:
\begin{minted}{fortran}
do i = 1, n
  b(n) = (a(i) + a(i+1)) / 2.0
\end{minted}
\item Most loop-array computations of the previous form read
from arrays with a pattern that includes the immediate
neighbours (offsets of 1 from the induction variables);
We grouped the data based on subclassifications of \emph{grouping 1}
in terms of contiguity and positioning.
%Focussing on grouping 1.
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
\textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
&\textbf{Position} & \textbf{Number} & \% {(of whole)} \\ \hline
Affine, contiguous
& 31 & 0.03\% \\ \hline
%\multirow{3}{*}{Affine, contiguous}
% & %\multirow{3}{*}{31}
% & %\multirow{3}{*}{0.03\%}
% & \textsf{away} & 4 & 0.00\% \\
%& 31 & 0.03\% & \textsf{origin} & 15 & 0.01\% \\
%&&& \textsf{straddle} & 12 & 0.01\% \\ \hline
Affine, non-contiguous
 & 3
 & 0.00\% & \\ \hline
\multirow{3}{*}{Neighbour, contiguous}
 & %\multirow{3}{*}{83689}
 & %\multirow{3}{*}{83.08\%}
 & \textsf{away} & 152 & 0.14\% \\
& 83689 & 76.94\% & \textsf{origin} & 81309 & 74.75\% \\
&&& \textsf{straddle} & 2228 & 2.05\% \\ \hline
Neighbour, non-contiguous
 & 4854
 & 4.46\% \\ \hline
Other & 20196 & 18.57\%  \\
 \hline \hline
\textbf{Total}  & 88577    &  \\
Thus, $76.94\%$ of array computations have only contiguous neighbour
indices (which may include constants in some indices) in a contiguous
pattern on the right hand side, with $76.80\%$ either at the origin
($74.75\%$) or offset by a distance of 1 from the origin
($2.05\%$). This confirms both hypotheses 2 and 3. Only 152 instances
of index patterns not immediately next to or over the origin were
found.  There is a non trivial amount of non-contiguous arrays access,
but it is a much smaller proportion of the whole.

@@

and whose left-hand side is an neighbourhood-indexed array subscript
(\eg{}, the classic five-point stencil);

@@

included an orthogonal classification of whether
left-hand and right-hand sides are consistent (have a common subset
of induction variables) or not.

@@

& \textbf{\# consistent}

@@

affine or a constant translation. Thus, the data support ours

@@

\begin{minipage}{0.48\linewidth}

@@

Most stencils are contiguous, but there is a non trivial

@@

are non-trivial amounts of other kinds of relationship.

@@

\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
RHS neighbour/affine unique indices      & 78349     & 72.03\%  \\\hline
RHS neighbour/affine repeated indices   & 10228     & 9.40\%  \\\hline
other                             & 20196     & 18.57\%  \\\hline\hline
Total                             & 108773    &
\end{tabular}
\end{center}
%
Thus, hypothesis 5 is confirmed, the majority of array computations
have access patterns with linear usage of array subscripts.

>>>>

\item Many array computations are \emph{stencil computations}: an
  assignment whose right-hand side comprises array subscripts with
  neighbourhood indices and whose left-hand side is a
  neighbourhood-indexed array subscript (\eg{}, the Laplace operator);

@@

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we categorise the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l|l}
  Shorthand & Classification & Example \\ \hline
  (\textsf{const}) & Consistent & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  (\textsf{perm}) & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  (\textsf{LHSsub}) & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  (\textsf{LHSsup}) & LHS superset & \fortran{a(i, j) = b(i)} \\
  (\textsf{inconst}) & Inconsistent & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%
If an assignment is not inconsistent, but one of the other four
categories, we refer to it as *consistent.

@@

\noindent
We revisit each hypothesis and show the related data and conclusions.
%
\begin{enumerate}
\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or dervied) induction variables;
\end{enumerate}
%
We identified $108,773$ instances of assignments within loops in
which an array subscript flows to the right-hand side. We refer
to each one of these as an \emph{array computation}.
We performed the classification described above
and grouped the data by the right-hand side classification only:
%%
\begin{center}
\begin{tabular}{l|S[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
\textbf{RHS classification} & \textbf{Number} & \% &
\textit{grouping 1} \\ \hline
Affine                          & 34        & 0.03\%
& \multirow{3}{*}{81.43\%} \\ \cline{1-3}
Neighbourhood                   & 65334     & 60.06 \%  \\ \cline{1-3}
Neighbourhood + absolute       & 23209     & 21.34\%  \\ \hline
Other                           & 20196     & 18.57 \%  \\ \hline \hline
\textbf{Total}                  & 108773    & 100\% \\
\end{tabular}
\end{center}
%
\noindent
Since, affine and neighbour patterns are static, we group them
(\textit{grouping 1}) to see that $81.43\%$ of the array computations
(88,577) have a static pattern that comprises constant offsets of
base/derived induction variables (potential with some absolute
indices), confirming \emph{hypothesis 1}. Note that affine subscripts
are rare, perhaps due to programmers relying on compilers for
optimisation rather than hand optimising (which tends to lead to
affine indices).  Overwhelmingly, the main category is neighbourhood
offsets (possibly mixed with constants in some dimensions). The
``affine + absolute'' class is not represented at all. \marginpar{\mnote{(2)
  Mention what sort of things are included in other.}}

\begin{enumerate}[resume]
\item Most loop-array computations of the previous form read
from arrays with a \emph{contiguous} pattern.
%
\item Most loop-array computations of the previous form read
from arrays with a pattern that includes the immediate
neighbours (offsets of 1 from the induction variables);
\end{enumerate}
%
We grouped the data based on subclassifications of \emph{grouping 1}
in terms of contiguity and positioning.
%Focussing on grouping 1.
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
\textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
&\textbf{Position} & \textbf{Number} & \% {(of whole)} \\ \hline
Affine, contiguous
& 31 & 0.03\% \\ \hline
%\multirow{3}{*}{Affine, contiguous}
% & %\multirow{3}{*}{31}
% & %\multirow{3}{*}{0.03\%}
% & \textsf{away} & 4 & 0.00\% \\
%& 31 & 0.03\% & \textsf{origin} & 15 & 0.01\% \\
%&&& \textsf{straddle} & 12 & 0.01\% \\ \hline
Affine, non-contiguous
 & 3
 & 0.00\% & \\ \hline
\multirow{3}{*}{Neighbour, contiguous}
 & %\multirow{3}{*}{83689}
 & %\multirow{3}{*}{83.08\%}
 & \textsf{away} & 152 & 0.14\% \\
& 83689 & 76.94\% & \textsf{origin} & 81309 & 74.75\% \\
&&& \textsf{straddle} & 2228 & 2.05\% \\ \hline
Neighbour, non-contiguous
 & 4854
 & 4.46\% \\ \hline
Other & 20196 & 18.57\%  \\
 \hline \hline
\textbf{Total}  & 108773    & 100\% \\
\end{tabular}
\end{center}
%%
Thus, $76.94\%$ of array computations have only neighbour
indices (which may include some absolute indices) in a contiguous
pattern on the right hand side, with $76.80\%$ either including the origin
($74.75\%$) or offset by a distance of 1 from the origin
($2.05\%$). This confirms both hypotheses 2 and 3. Only 152 instances
of index patterns not immediately next to or over the origin were
found.  There is a non trivial amount of non-contiguous arrays access,
but it is a much smaller proportion of the whole.

@@

and whose left-hand side is a neighbourhood array subscript.

@@

included the orthogonal classification of whether
left- and right-hand sides are inconsistent or
otherwise (have a common subset of induction variables).

@@

& \textbf{\# *consistent}

@@

affine or a constant translation. Thus, the data supports ours

@@

\begin{minipage}{0.42\linewidth}
\hspace{-1.6em}

@@

Most stencils are contiguous, but there is a non-trivial

@@

are non-trivial amounts of the other kinds of relationship.

@@

\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
RHS neighbour/affine unique subscripts      & 78349     & 72.03\%  \\\hline
RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
other                             & 20196     & 18.57\%  \\\hline\hline
Total                             & 108773    &
\end{tabular}
\end{center}
%
Thus, hypothesis 5 is confirmed, the majority of array computations
have access patterns with linear usage of array subscripts.
